---
title: "Getting started with richcast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with richcast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`richcast` answers one question: **how did the number of species in a place
change over time, as climate changed?**

It gets there by fitting a distribution model per species against present-day
climate -- an equally weighted ensemble of a random forest and MaxEnt, trained
on pseudo-presences sampled from the species' IUCN range polygon -- projecting
each model onto a series of palaeoclimate reconstructions, and stacking the
results into a richness surface for each slice.

This vignette walks through the API on synthetic data, so every chunk here
actually runs.

```{r setup}
library(richcast)
```

## The four objects

Nearly everything in `richcast` is one of four things.

| Object | Built by | Holds |
|---|---|---|
| range source | `iucn_folder()`, `iucn_shapefile()`, `sf_polygons()` | where range polygons come from |
| taxon database | `build_taxon_db()` | species, ranges, traits |
| climate source | `climate_dir()`, `pastclim_climate()` | how to fetch a slice |
| region | `region()` | the area to model richness over |

With real data, the range source is usually a folder of Red List downloads:

```{r eval = FALSE}
db <- build_taxon_db(iucn_folder("UngulatePolygons"))
```

`iucn_folder()` reads every shapefile under the folder, keeps only polygons
coded as extant and native or reintroduced, and dissolves each species into a
single range.

## A taxon database

Ranges come from you. `richcast` bundles no range data: IUCN Red List polygons
cannot legally be redistributed, and everyone's taxon is different anyway.
Here is a two-species toy set, one species deliberately split across two
disjunct patches the way real range databases store things.

```{r}
square <- function(xmin, ymin, size) {
  sf::st_polygon(list(cbind(
    c(xmin, xmin + size, xmin + size, xmin, xmin),
    c(ymin, ymin, ymin + size, ymin + size, ymin)
  )))
}

ranges <- sf::st_sf(
  species = c("Genus_alpha", "Genus_alpha", "Genus_beta"),
  geometry = sf::st_sfc(
    square(0.5, 0.5, 3), square(6, 0.5, 2), square(5, 5, 4),
    crs = 4326
  )
)

db <- build_taxon_db(sf_polygons(ranges))
db
```

Note the message: three features became two species. This matters more than it
looks. Treating each feature as a species means a model trains on one fragment
while the *rest of that species' range* is handed to the background sample --
so the model is asked to tell the species apart from itself. Pass
`dissolve = FALSE` if you genuinely want per-feature modelling.

### Joining traits

`build_taxon_db()` takes an optional trait table, joined on a normalised
species name so that `"Genus alpha"`, `"Genus_alpha"` and `"genus alpha"` all
land on the same row.

```{r}
traits <- data.frame(
  species = c("Genus alpha", "Genus beta"),
  mass    = c(20, 400)
)

db <- build_taxon_db(sf_polygons(ranges), traits = traits)
filter_taxa(db, mass < 100)
```

Both functions report what they did. Trait joins and filters are where species
quietly disappear, so the counts are printed rather than left for you to
discover later.

## A climate source

For real work, prepare slices once with `prepare_climate()`, which streams
through `pastclim`, aggregates, and writes one GeoTIFF per variable per slice:

```{r eval = FALSE}
clim <- prepare_climate(
  path   = "climate/eurasia",
  vars   = c("bio01", "bio12"),
  times  = seq(850, 1850, by = 100),
  extent = c(-15, 180, 10, 82)
)
```

It skips files that already exist, so an interrupted download resumes. Point
`climate_dir()` at the result afterwards.

For this vignette we fabricate a climate gradient instead.

```{r}
clim_dir <- file.path(tempdir(), "toy_climate")

make_slice <- function(label, shift) {
  d <- file.path(clim_dir, label)
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
  r <- terra::rast(nrows = 40, ncols = 40, xmin = 0, xmax = 10,
                   ymin = 0, ymax = 10, crs = "EPSG:4326")
  xy <- terra::xyFromCell(r, seq_len(terra::ncell(r)))
  terra::values(r) <- xy[, 1] * 2 + xy[, 2] + shift
  terra::writeRaster(r, file.path(d, "bio01.tif"), overwrite = TRUE)
  terra::values(r) <- (10 - xy[, 1]) * 2 + xy[, 2] - shift
  terra::writeRaster(r, file.path(d, "bio12.tif"), overwrite = TRUE)
}

make_slice("present_1985", 0)
for (i in seq_along(times <- c(850, 950, 1050, 1150))) {
  make_slice(sprintf("time_%04d", times[i]), i * 0.8)
}

clim <- climate_dir(clim_dir)
clim
```

Every slice must share a grid. A mismatch between the grid a model was fitted
on and the grid it is projected onto yields a suitability surface that looks
entirely plausible and means nothing, so check rather than assume:

```{r}
check_climate_grids(clim, vars = c("bio01", "bio12"))
```

## Fitting once, projecting many times

A model depends only on present-day climate and the species' range, so it is
fitted **once** and reused for every slice.

```{r}
land <- sf::st_sfc(square(-1, -1, 12), crs = 4326)

m <- fit_sdm(db, "Genus_alpha", clim,
             predictors = c("bio01", "bio12"),
             land = land, quiet = TRUE)
m
```

The study extent is the range's bounding box widened by 30% of its diagonal.
Inside it, 100 pseudo-presences are drawn from the range polygon and 1000
pseudo-absences from the rest of the extent. The random forest is fitted to
those points; MaxEnt contrasts the pseudo-presences with background drawn
from the whole extent, inside the range as well (`maxnet_background`). Their
predictions are averaged. The tenth
percentile of ensemble suitability at the presences (`p10`) sets the
presence threshold.

`m$metrics` scores each member and the ensemble on a 25% hold-out, by AUC and
by the continuous Boyce index:

```{r}
m$metrics
```

`land` masks predictions to land and keeps background sampling off the ocean.
It defaults to Natural Earth; here the toy world is entirely land.

Projection is a separate call, so the cost of adding a time slice is one
prediction, not a refit:

```{r}
for (tt in times) {
  p <- project_sdm(m, clim, tt, quiet = TRUE)
  cat(sprintf("%5d CE: %4d suitable cells\n", tt, p$cells))
}
```

Both objects survive `saveRDS()`. Bare `terra` objects store external pointers
that reload as invalid handles, so `richcast` keeps vectors as `sf` and rasters
wrapped.

```{r}
f <- tempfile(fileext = ".rds")
saveRDS(m, f)
suitability(readRDS(f))
```

## Regions

A region is either a preset -- `"europe"`, `"asia"`, `"middle_east"`,
`"africa"`, `"north_america"` or `"south_america"` -- or your own box:

```{r}
region("middle_east")
toy <- region(c(0, 10, 0, 10), label = "toy region")
```

When hindcasting a region, only species whose present-day range lies within
10 degrees of it are trained. The rule is fixed and always applied, and
`species_near()` shows which species it keeps.

```{r}
species_near(db, toy)
```

## The whole thing at once

`run_hindcast_series()` fits every listed species near the region, projects
each onto every slice, and stacks the results. For a preset region the list
defaults to the species reported from that region's Pleistocene
zooarchaeological record -- see `zooarch_taxa()` -- and a custom box like this
one needs its own list.

```{r}
res <- run_hindcast_series(
  db, clim,
  times      = times,
  region     = toy,
  species    = c("Genus_alpha", "Genus_beta"),
  predictors = c("bio01", "bio12"),
  land       = land,
  quiet      = TRUE
)
res
```

Richness per slice. `mean_richness` counts species above their threshold.
(`mean_expected`, the summed suitability, is deprecated: suitabilities are not
calibrated probabilities, so their sum is not a richness estimate.)

```{r}
res$richness
```

Model metrics, per member and for the ensemble:

```{r}
res$models
```

Per-species change, over each species' own study extent (`cells`) and over
the region (`region_cells`):

```{r}
res$species
```

The surfaces themselves are kept, as `richness_surface()` rasters or as a tidy
`richness_grid()` table ready for plotting:

```{r}
richness_surface(res, 1050)
```

## Richness at a point

`richness_at()` evaluates every model at a coordinate. It returns the richness
there and the species expected, ranked by suitability:

```{r}
pt <- richness_at(res, lon = 2, lat = 2, time = c("present", 1050),
                  climate = clim)
pt
pt$species
```

Presence uses each model's p10 threshold by default. Every model also stores
the cutoff that maximises the true skill statistic, which is usually more
lenient on a wide study extent; `threshold = "tss"` applies it instead, and
the two counts bracket a plausible richness:

```{r}
richness_at(res, lon = 2, lat = 2, time = c("present", 1050),
            climate = clim, threshold = "tss")$richness
```

## Where to go next

* `?fit_sdm` -- sampling, the ensemble and the metrics.
* `?iucn_folder` -- which IUCN polygons count as today's range.
* `?prepare_climate` -- getting palaeoclimate onto disk.
